Catastrophic Goodhart: regularizing RLHF with KL divergence does not mitigate heavy-tailed reward misspecification

### 2024-10-24 paper/main.tex
DMRMDP trajectories must terminate in a sink state
combine theorems 1 and 2 (previously 3) into equivalence statement

When applying KL regularization, the trained model is regularized towards some base policy $\pi_0$. One would hope that a KL penalty can produce good outcomes even in the case of reward misspecification; that is, if the reward $U$ is the sum of true utility $V$ and an error term $X$, we would hope that optimal policies under a KL penalty achieve high $V$ even if the magnitude of $X$ is large. We show that this is not always the case: Corollary~\ref{cor:heavy-tailed} of Theorems~\ref{thm1}, \ref{thm2}, and~\ref{thm3} establishes that when $X(\pi_0)$ is heavy-tailed, there are arbitrarily well-performing policies $\pi$ with $\mathbb E_{\pi}[V] \approx \mathbb E_{\pi_0}[V]$. However, Theorem~\ref{thm4} shows that when error is light-tailed and independent of $V$, the optimal policy under a KL penalty results in $V > 0$, and $V$ can be made arbitrarily large. Thus, the tails of the error distribution are crucial in determining how much utility will result from optimization towards an imperfect proxy.

@@

@@

\begin{itemize}
\item
  The transition function
  \(P: \mathcal{S} \times \mathcal{A} \to \mathcal{S}\) is
  deterministic, i.e., for each state \(s \in \mathcal{S}\) and action
  \(a \in \mathcal{A}\), there exists a unique state
  \(s' \in \mathcal{S}\) such that \(P(s' | s, a) = 1\). \\ \textbf{In RLHF:} the transition is appending the generated token $a$ to the context $s$.
\item
  There is a set of sink states \(E \subseteq \mathcal S\) that
  terminate a trajectory, which is disjoint from the set of start
  states. \\ \textbf{In RLHF:} The sink states are sequences ending in \texttt{<EOS>} or above a certain length.
\item
  Returns are Markovian; that is, for any two trajectories
  \(\tau=(s_1, a_1, \dots, s_n), \tau'=(s_1', a_1', \dots, s_n'),\) if
  \(s_{n}= s'_{n}\), then \(\tau\) and \(\tau'\) have identical return
  distributions. Equivalently, for the trajectory random variable
  \(T=(S_1, A_1, \dots)\) distributed according to any policy, with
  return \(G\), \(G \bot\!\!\!\!\!\ \bot (S_{<i}, A_{<i}) \ |\ S_i\) for
  any \(i \ge 1\). \\ \textbf{In RLHF:} the return only depends on the full generated string, which is the final state.
\end{itemize}

@@

\begin{theorem}
\label{thm2}
Let \(W = (\mathcal S, \mathcal A, P, R)\) be
a deterministic-transition MDP with Markovian returns. Given \(W\) we
define the function that takes policies to trajectories
\(Tr: (S \to \Delta A) \to \Delta(S \times A)^*\), and the average
return function \(g: (S \times A)^* \to \mathbb R\), which induces a
function \(G: \Delta(S \times A)^* \to \Delta \mathbb R\). Let
\(\pi_0: \mathcal S \to \Delta \mathcal A\) be some base policy. If
\(G \circ Tr(\pi_0)\) is heavy-tailed with finite mean \(\mu_Q\), then
for any \(M, \epsilon > 0\), there is a policy \(\pi\) with mean return
\(\mathbb E[U | U \sim G \circ Tr(\pi)] > M\) and
\(\mathbb E_{s \in T, T\sim Tr(\pi)}[D_{KL}(\pi(s) \| \pi_0(s))] < \epsilon\).
\end{theorem}

@@

\begin{theorem}
\label{thm3} If \(V\) is light-tailed and \(d = D_{KL}(P \| Q)\) is bounded, then
\(\mathbb E_P[V]\) is bounded, and \(\mathbb E_P[V] - \mathbb E_Q[V] \to 0\) as
\(d \to 0\).
\end{theorem}

@@

When $\alpha$ is small but much larger than $q(x)$, we approximate this to first order as $D_{KL}(P \| Q) \approx \alpha \log \left(\frac{\alpha}{q(x)}\right)$. In Theorems 1 and 2, we prove that when the error is sufficiently heavy-tailed, a policy that gets extremely large reward a small fraction of the time will achieve high expected reward with low KL divergence. This is not the case here because the rewards achieved through ACG were small and the log-probabilities extremely negative. For example, a policy that matches Llama 2-chat's base reward 99\% of the time and uses the highest-reward input generated by ACG $\alpha=$1\% of the time will have KL divergence from Llama 2-chat of $\alpha(\log(\alpha) - 1339.70) = 13.35$ nats, but reward only about $\alpha*(2.2377-0.3329) = 0.02571$ greater than the base model, far less than can be obtained with the same KL divergence by conditioning.

@@

\begin{itemize}
    \item If we maximize $\sigma(\mathbb E[U]) + D_{KL}(Tr(\pi) \| Tr(\pi_0))$, where $\sigma$ is a bounded function (e.g. sigmoid), all near-optimal policies will have $V \approx 0$. Since we can only obtain so much reward from $\sigma(\mathbb{E}[U])$, it pays to make the KL (and thus V) go to zero.
    \item If we cap KL to a finite value (or dynamically adjust the KL penalty to target a finite KL, as done in \citet{ziegler2020finetuning}, then $\mathbb E[V]$ is also upper bounded by a finite value (see Theorem 3), and we think it is likely that $\mathbb E[V] \approx 0$. Consider a toy model where an AI can adjust three parameters: true quality $V$ of responses, frequency of reward hacking (producing actions with extremely high X), and severity of hacking (value of X on those actions). Adjusting the policy to increase $\mathbb E[U]$ without increasing KL increase the severity of hacking while decreasing either frequency of hacking or quality of responses. When $E[U]$ is already large, decreasing quality has much better returns than decreasing frequency. This is similar to Theorems \ref{thm5}, \ref{thm6} about hard-threshold optimization.
    \item Any way we maximize $\mathbb{E} [U(\pi)]-\beta D_{K L}\left(\pi, \pi_0 \right)$ results in very large values of $\mathbb E[U(\pi)]$, and there are a number of arguments that extreme optimization for an imperfect proxy can result in decreased utility due to tradeoffs between $X$ and $V$; e.g., the constrained resource scenario in \citep{zhuang2021consequences}.
\end{itemize}

@@

Theorems 2 and 3 do not require any independence assumption, but Theorems 4, 5, and 6 require that error $X$ and utility $V$ are independent, which seems to be violated in practice. Future work could weaken this assumption, although intuitively obvious ways to weaken it result in the statement being false. \footnote{Suppose that error $X$ is light-tailed conditional on any value of $V$, but our proxy is merely unbiased ($\mathbb E[X|V=v]=0$ for all $v$). Then the limit of $V$ under optimization for $X+V$ still depends on the relationship between $X$ and $V$. If they are independent, Theorem 6 says that \(\lim_{t \to\infty} \mathbb E[V | X + V \ge t] = \infty\). But if \(V \sim N(0, 1)\), and \(X | V \sim N(0, 4)\) when \(V \in [-1, 1]\), otherwise \(X=0\), then \(\lim_{t \to\infty} \mathbb E[V | X + V \ge t] = 0\).}

@@

\subsection{Theorem 2}
\newtheorem*{theorem2}{Restatement of Theorem \ref{thm2}}
\begin{theorem2}
Let \(W = (\mathcal S, \mathcal A, P, R)\) be
a deterministic-transition MDP with Markovian returns. Given \(W\), we
define the function that takes policies to trajectories
\(Tr: (S \to \Delta A) \to \Delta(S \times A)^*\), and the average
return function \(g: (S \times A)^* \to \mathbb R\) which induces a
function \(G: \Delta(S \times A)^* \to \Delta \mathbb R\). Let
\(\pi_0: \mathcal S \to \Delta \mathcal A\) be some base policy. If
\(G \circ Tr(\pi_0)\) is heavy-tailed with finite mean \(\mu_Q\), then
for any \(M, \epsilon > 0\), there is a policy \(\pi\) with mean return
\(\mathbb E[U | U \sim G \circ Tr(\pi)] > M\) and
\(\mathbb E_{s \in T, T\sim Tr(\pi)}[D_{KL}(\pi(s) \| \pi_0(s))] < \epsilon\).
\end{theorem2}

@@

\subsection{Theorem 3}
\newtheorem*{theorem3}{Restatement of Theorem \ref{thm3}}
\begin{theorem3} If \(V\) is light-tailed, \(\mathbb E_Q[V]\)
is zero, and \(d = D_{KL}(P \| Q)\) is bounded, then
\(\mathbb E_P[V]\) is bounded, and \(\mathbb E_P[V] \to 0\) as
\(d \to 0\).
\end{theorem3}

>>>>

When applying KL regularization, the trained model is regularized towards some base policy $\pi_0$. One would hope that a KL penalty can produce good outcomes even in the case of reward misspecification; that is, if the reward $U$ is the sum of true utility $V$ and an error term $X$, we would hope that optimal policies under a KL penalty achieve high $V$ even if the magnitude of $X$ is large. We show that this is not always the case: Corollary~\ref{cor:heavy-tailed} of Theorems~\ref{thm1}, \ref{thm3}, and~\ref{thm2} establishes that when $X(\pi_0)$ is heavy-tailed, there are arbitrarily well-performing policies $\pi$ with $\mathbb E_{\pi}[V] \approx \mathbb E_{\pi_0}[V]$. However, Theorem~\ref{thm4} shows that when error is light-tailed and independent of $V$, the optimal policy under a KL penalty results in $V > 0$, and $V$ can be made arbitrarily large. Thus, the tails of the error distribution are crucial in determining how much utility will result from optimization towards an imperfect proxy.

@@

\begin{theorem}
    \label{thm2} However, if \(V\) is light-tailed and \(d = D_{KL}(P \| Q)\) is bounded, then
    \(\mathbb E_P[V]\) is bounded, and \(\mathbb E_P[V] - \mathbb E_Q[V] \to 0\) as
    \(d \to 0\).
    \end{theorem}

@@

\begin{itemize}
\item
  The transition function
  \(P: \mathcal{S} \times \mathcal{A} \to \mathcal{S}\) is
  deterministic, i.e., for each state \(s \in \mathcal{S}\) and action
  \(a \in \mathcal{A}\), there exists a unique state
  \(s' \in \mathcal{S}\) such that \(P(s' | s, a) = 1\). \\ \textbf{In RLHF:} the transition is appending the generated token $a$ to the context $s$.
\item
  There is a set of sink states \(E \subseteq \mathcal S\) that
  terminate every trajectory, which is disjoint from the set of start
  states. \\ \textbf{In RLHF:} The sink states are sequences ending in \texttt{<EOS>} or above a certain length.
\item
  Returns are Markovian; that is, for any two trajectories
  \(\tau=(s_1, a_1, \dots, s_n), \tau'=(s_1', a_1', \dots, s_n'),\) if
  \(s_{n}= s'_{n}\), then \(\tau\) and \(\tau'\) have identical return
  distributions. Equivalently, for the trajectory random variable
  \(T=(S_1, A_1, \dots)\) distributed according to any policy, with
  return \(G\), \(G \bot\!\!\!\!\!\ \bot (S_{<i}, A_{<i}) \ |\ S_i\) for
  any \(i \ge 1\). \\ \textbf{In RLHF:} the return only depends on the full generated string, which is the final state.
\end{itemize}

@@

\begin{theorem}
\label{thm3}
Let \(W = (\mathcal S, \mathcal A, P, R)\) be
a deterministic-transition MDP with Markovian returns. Given \(W\) we
define the function that takes policies to trajectories
\(Tr: (S \to \Delta A) \to \Delta(S \times A)^*\), and the average
return function \(g: (S \times A)^* \to \mathbb R\), which induces a
function \(G: \Delta(S \times A)^* \to \Delta \mathbb R\). Let
\(\pi_0: \mathcal S \to \Delta \mathcal A\) be some base policy. If
\(G \circ Tr(\pi_0)\) is heavy-tailed with finite mean \(\mu_Q\), then
for any \(M, \epsilon > 0\), there is a policy \(\pi\) with mean return
\(\mathbb E[U | U \sim G \circ Tr(\pi)] > M\) and
\(\mathbb E_{s \in T, T\sim Tr(\pi)}[D_{KL}(\pi(s) \| \pi_0(s))] < \epsilon\).
\end{theorem}

@@

@@

When $\alpha$ is small but much larger than $q(x)$, we approximate this to first order as $D_{KL}(P \| Q) \approx \alpha \log \left(\frac{\alpha}{q(x)}\right)$. In Theorems 1 and 3, we prove that when the error is sufficiently heavy-tailed, a policy that gets extremely large reward a small fraction of the time will achieve high expected reward with low KL divergence. This is not the case here because the rewards achieved through ACG were small and the log-probabilities extremely negative. For example, a policy that matches Llama 2-chat's base reward 99\% of the time and uses the highest-reward input generated by ACG $\alpha=$1\% of the time will have KL divergence from Llama 2-chat of $\alpha(\log(\alpha) - 1339.70) = 13.35$ nats, but reward only about $\alpha*(2.2377-0.3329) = 0.02571$ greater than the base model, far less than can be obtained with the same KL divergence by conditioning.

@@

\begin{itemize}
    \item If we maximize $\sigma(\mathbb E[U]) + D_{KL}(Tr(\pi) \| Tr(\pi_0))$, where $\sigma$ is a bounded function (e.g. sigmoid), all near-optimal policies will have $V \approx 0$. Since we can only obtain so much reward from $\sigma(\mathbb{E}[U])$, it pays to make the KL (and thus V) go to zero.
    \item If we cap KL to a finite value (or dynamically adjust the KL penalty to target a finite KL, as done in \citet{ziegler2020finetuning}, then $\mathbb E[V]$ is also upper bounded by a finite value (see Theorem \ref{thm2}), and we think it is likely that $\mathbb E[V] \approx 0$. Consider a toy model where an AI can adjust three parameters: true quality $V$ of responses, frequency of reward hacking (producing actions with extremely high X), and severity of hacking (value of X on those actions). Adjusting the policy to increase $\mathbb E[U]$ without increasing KL increase the severity of hacking while decreasing either frequency of hacking or quality of responses. When $E[U]$ is already large, decreasing quality has much better returns than decreasing frequency. This is similar to Theorems \ref{thm5}, \ref{thm6} about hard-threshold optimization.
    \item Any way we maximize $\mathbb{E} [U(\pi)]-\beta D_{K L}\left(\pi, \pi_0 \right)$ results in very large values of $\mathbb E[U(\pi)]$, and there are a number of arguments that extreme optimization for an imperfect proxy can result in decreased utility due to tradeoffs between $X$ and $V$; e.g., the constrained resource scenario in \citep{zhuang2021consequences}.
\end{itemize}

@@

Theorems 1-3 do not require any independence assumption, but Theorems \ref{thm4}, \ref{thm5}, and \ref{thm6} require that error $X$ and utility $V$ are independent, which seems to be violated in practice. Future work could weaken this assumption, although intuitively obvious ways to weaken it result in the statement being false. \footnote{Suppose that error $X$ is light-tailed conditional on any value of $V$, but our proxy is merely unbiased ($\mathbb E[X|V=v]=0$ for all $v$). Then the limit of $V$ under optimization for $X+V$ still depends on the relationship between $X$ and $V$. If they are independent, Theorem 6 says that \(\lim_{t \to\infty} \mathbb E[V | X + V \ge t] = \infty\). But if \(V \sim N(0, 1)\), and \(X | V \sim N(0, 4)\) when \(V \in [-1, 1]\), otherwise \(X=0\), then \(\lim_{t \to\infty} \mathbb E[V | X + V \ge t] = 0\).}

@@

\subsection{Theorem 2}
\newtheorem*{theorem2}{Restatement of Theorem \ref{thm2}}
\begin{theorem2} If \(V\) is light-tailed, \(\mathbb E_Q[V]\)
is zero, and \(d = D_{KL}(P \| Q)\) is bounded, then
\(\mathbb E_P[V]\) is bounded, and \(\mathbb E_P[V] \to 0\) as
\(d \to 0\).
\end{theorem2}

\subsection{Theorem 3}
\newtheorem*{theorem3}{Restatement of Theorem \ref{thm3}}
\begin{theorem3}
Let \(W = (\mathcal S, \mathcal A, P, R)\) be
a deterministic-transition MDP with Markovian returns. Given \(W\), we
define the function that takes policies to trajectories
\(Tr: (S \to \Delta A) \to \Delta(S \times A)^*\), and the average
return function \(g: (S \times A)^* \to \mathbb R\) which induces a
function \(G: \Delta(S \times A)^* \to \Delta \mathbb R\). Let
\(\pi_0: \mathcal S \to \Delta \mathcal A\) be some base policy. If
\(G \circ Tr(\pi_0)\) is heavy-tailed with finite mean \(\mu_Q\), then
for any \(M, \epsilon > 0\), there is a policy \(\pi\) with mean return
\(\mathbb E[U | U \sim G \circ Tr(\pi)] > M\) and
\(\mathbb E_{s \in T, T\sim Tr(\pi)}[D_{KL}(\pi(s) \| \pi_0(s))] < \epsilon\).
\end{theorem3}

@@
